% =============================================================================
%  Programmatic-life decomposition — skeleton paper
%  STATUS: UNPROVEN. Worked-example paper exhibiting Decomp on Tierra.
% =============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\title{Programmatic Life as a Theorem Instance: \\
       Decomposition of Tierra under $\mathrm{Decomp}$ \\
       \large (UNPROVEN --- skeleton paper)}
\author{Liam McGuire\thanks{Unaffiliated researcher, Seattle, USA. liammcg44@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper is a \textbf{skeleton} for a worked-example paper that
specialises the decomposition functor $\mathrm{Decomp}$
(\texttt{n1\_necessity/}) on an active Tierra arms-race fragment,
exhibits the four slots concretely, and reports vitality profile
$(2, (1, 1))$. \textbf{No proof is attempted.} The exhibition
requires N1's Decomp construction to be complete first. The result is
UNPROVEN.
\end{abstract}

\section{Goal (UNPROVEN)}

Exhibit, for a specific Tierra-arms-race fragment, the four-slot
system $(\X, V, F, T) \in \mathbf{Sys}$ that the decomposition
functor $\mathrm{Decomp}$ produces, and verify that its vitality
profile is $(2, (1, 1))$. This is the operational realisation of the
framework's predictive thesis: programmatic life and biological life
at the same vitality profile become the same categorical object
(\texttt{cat1\_characterisation/}).

\section{Plan of attack (no proof attempted)}

\begin{enumerate}[leftmargin=2em,itemsep=4pt]
  \item \textbf{Choose the fragment (UNPROVEN).} Pick a specific
    Tierra run with host, parasite, and hyperparasite lineages
    active. Bound the configuration: fixed memory size, fixed
    instruction set, fixed mutation rate.
  \item \textbf{Present as $\mathcal{O}_W$-algebra (UNPROVEN).}
    Encode the simulation as an algebra of wiring diagrams; the
    boxes are programs, the wires are memory-address dependencies.
  \item \textbf{Apply $\mathrm{Decomp}$ (UNPROVEN; needs N1).}
    Exhibit the four-slot system $(\X, V, F, T)$ produced by the
    functor. Compare against the descriptive Tierra slot reading
    given in the scaffolding paper.
  \item \textbf{Compute vitality profile (UNPROVEN; needs
    \texttt{vitality\_profile\_methodology/}).} Apply the algorithm
    of \texttt{vitality\_computation.md} and verify $(k, \sigma) =
    (2, (1, 1))$ during the arms-race phase, $(2, (1, 0))$ after
    stabilisation.
\end{enumerate}

\section{Status and follow-up}

\textbf{Current status: UNPROVEN.} Depends on N1 and on the
methodology paper. Once both land, this paper is the cheapest
empirical demonstration that the framework's predictive thesis is
operationally cashed out.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
