% =============================================================================
%  L1 (= renamed OP1): Causal Emergence of Game of Life
%      under Markov-blanket viability — skeleton paper
%  STATUS: UNPROVEN. This file is a skeleton; the proof is not attempted here.
% =============================================================================
\documentclass[11pt]{article}

% ---------- Encoding & fonts ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}

% ---------- Math ----------
\usepackage{amsmath, amssymb, amsthm, mathtools}

% ---------- Layout ----------
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}

% ---------- Hyperlinks ----------
\usepackage[hidelinks]{hyperref}

% ---------- Theorem environments ----------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem*{namedconjecture}{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

% ---------- Math shortcuts (mirror scaffolding paper) ----------
\DeclareMathOperator{\EI}{EI}
\DeclareMathOperator{\CE}{CE}
\newcommand{\X}{\mathbf{X}}

% =============================================================================
\title{Causal Emergence of Game of Life under \\
       Markov-blanket Viability \\
       \large (UNPROVEN --- skeleton paper for L1)}
\author{Liam McGuire\thanks{Unaffiliated researcher, Seattle, USA. liammcg44@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper is a \textbf{skeleton} that records the statement of L1
(= renamed OP1) from the scaffolding paper~\cite{emergent_systems_2026},
identifies the lowest-cost candidate witness system
(Game of Life under Markov-blanket viability), and sketches the
proof-attack plan. \textbf{No proof is attempted in this paper as of
its current revision.} The result is UNPROVEN. The paper exists so
that a future proof can be developed against a fixed statement and
plan, with status tracked in \texttt{RESULTS.md}.
\end{abstract}

% -----------------------------------------------------------------------------
\section{Statement (UNPROVEN)}

\begin{quote}
\emph{``L1 (= renamed OP1): causal emergence of scaffolded systems ---
Find substrate--viability pairs $(\X, F)$ for which one can prove
$\sup_\Pi \CE(M(\mathcal{S}), \Pi) > 0$ for the scaffolded system
$\mathcal{S}$.''}
\end{quote}

\noindent (Scaffolding paper~\cite{emergent_systems_2026},
\S\texttt{sec:op1}.)

This is L1's role in the framework's conjecture set: a positive
answer is a sufficiency witness for the top-level conjecture S1
($\mathbf{LifeCat}$ is non-empty), and is therefore the first
concrete result the framework targets.

% -----------------------------------------------------------------------------
\section{Result (target, UNPROVEN)}

The result this paper \emph{intends} to prove --- but does not yet ---
is the existence direction of L1 on a specific candidate:

\begin{quote}
\emph{Conjectured.} Game of Life with a Markov-blanket viability filter
$F_{\mathrm{MB}}$ on bounded-support configurations is causally
emergent in the sense of $\sup_\Pi \CE(M(\mathcal{S}), \Pi) > 0$.
\end{quote}

The candidate is taken from \texttt{proof\_techniques.md} §OP1.D,
``Lowest-hanging fruit.''

% -----------------------------------------------------------------------------
\section{Plan of attack (no proof attempted)}

The plan below is the proposed proof structure. \emph{No step is
discharged in this paper.} Each step has a UNPROVEN tag; future
revisions of this paper will discharge them one at a time.

\begin{enumerate}[leftmargin=2em,itemsep=4pt]
  \item \textbf{Finite-Markov reduction (UNPROVEN).} Restrict GoL to
    bounded-support configurations of total area $\leq A$. The
    resulting state space is finite; the GoL update is deterministic;
    the induced Markov chain $M(\mathcal{S})$ is finite-state. This is
    routine combinatorics and is included as the first lemma so the
    rest of the proof has a finite chain to work with.
  \item \textbf{Markov-blanket filter (UNPROVEN).} Construct
    $F_{\mathrm{MB}}$ on the finite state space by following the
    sparse-coupling-to-blanket-structure conditions of Heins \& Da Costa
    \cite{heins2022sparse}. Verify that the resulting filter is
    non-trivial (Definition non-trivial F in the scaffolding paper)
    and has discriminative capacity.
  \item \textbf{Macroscale partition $\Pi^*$ (UNPROVEN).} Promote the
    Israeli--Goldenfeld~\cite{israeli2004irreducibility,israeli2006coarse}
    block-coarse-grainings for related ECAs to GoL. Identify a
    candidate $\Pi^*$ partitioning configurations by their
    oscillator / glider / still-life equivalence classes.
  \item \textbf{Lumpability (UNPROVEN).} Verify the Kemeny--Snell
    strong-lumpability condition $U V P V = P V$ on
    $(M(\mathcal{S}), \Pi^*)$. Strong lumpability gives a
    well-defined macro Markov chain $M^{\Pi^*}$.
  \item \textbf{EI comparison (UNPROVEN).} Compute $\EI(M^{\Pi^*})$
    and $\EI(M(\mathcal{S}))$ via the
    Hoel--Albantakis--Tononi~\cite{hoel2013macro} determinism--degeneracy
    decomposition. Show that the macro-EI is strictly greater than
    the micro-EI on the bounded-support GoL with the chosen blanket
    filter, by bounding preimage multiplicity.
  \item \textbf{Robustness (UNPROVEN).} Apply
    Comolatti--Hoel~\cite{comolatti2022causal} to argue the
    $\CE > 0$ result is robust across causal measures (not an
    artefact of EI specifically).
\end{enumerate}

The above steps are listed as targets, not as proven lemmas. None of
them is discharged in this paper as of its current revision.

% -----------------------------------------------------------------------------
\section{Status and follow-up}

\textbf{Current status: UNPROVEN.} This paper currently exists as a
statement-and-plan skeleton, not as a proof. \texttt{RESULTS.md}
records L1 as \texttt{open}.

When the proof lands, the status should flip to \texttt{proved} and
this paper should expand the ``Plan of attack'' section into the
actual proof. A positive L1 result will simultaneously provide a
sufficiency witness for the framework's top-level conjecture S1; the
companion paper \texttt{s1\_sufficiency/} can then cite this paper
rather than constructing its own witness from scratch.

If a step in the plan fails (e.g., strong lumpability cannot be
verified for any natural $\Pi^*$ on GoL), the paper should record
the obstruction and pivot to one of the alternative candidate
systems flagged in the scaffolding paper §\texttt{sec:op1} ---
Flow-Lenia with Markov-blanket viability, Differentiable Logic CA,
POET-style coevolved environments --- with status
\texttt{mutated} in \texttt{RESULTS.md}.

% =============================================================================
\bibliographystyle{plain}
\bibliography{references}

\end{document}
