% =============================================================================
%  C3: Observer functor as divergence functional — skeleton paper
%  STATUS: UNPROVEN. Plausibly the easiest of the lemma-level conjectures.
% =============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem*{namedconjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\title{Observer Functor as Divergence Functional \\
       \large (UNPROVEN --- skeleton paper for C3)}
\author{Liam McGuire\thanks{Unaffiliated researcher, Seattle, USA. liammcg44@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper is a \textbf{skeleton} that records C3 from the scaffolding
paper~\cite{emergent_systems_2026}. \textbf{No proof is attempted.}
Plausibly the easiest of the lemma-level conjectures; the three
observer families very likely unify via a Reid--Williamson-style
divergence schema, but the proof is not written here.
The result is UNPROVEN.
\end{abstract}

\section{Statement (UNPROVEN)}

\begin{quote}
\emph{``C3 (observer functor as divergence functional): the external
observer functor $O : \mathbf{Sys} \to \mathbf{Obs}$ factors, in each
of the three families (density-distance, FM-embedding,
information-theoretic), through a single divergence-functional schema
$O(\mathcal{S})[\mu] = D(\mu \,\Vert\, \mu_{\mathrm{ref}})$ for a
context-dependent reference $\mu_{\mathrm{ref}}$.''}
\end{quote}
\noindent (Scaffolding paper~\cite{emergent_systems_2026},
\S\texttt{sec:c3}.)

\section{Plan of attack (no proof attempted)}

\begin{enumerate}[leftmargin=2em,itemsep=4pt]
  \item \textbf{Density-distance family (UNPROVEN, easy).} Fits by
    construction: novelty-search and Quality-Diversity observers
    score by a divergence against an archive or uniform reference.
    Citing Lehman \& Stanley~\cite{lehman2011novelty} and the QD
    literature is the work.
  \item \textbf{FM-embedding family (UNPROVEN, moderate).} ASAL and
    OMNI-EPIC observers fit when $\mu_{\mathrm{ref}}$ is taken as the
    archive's embedding distribution. The proof uses
    Belghazi et al.'s MINE construction and Sriperumbudur's IPM
    machinery to express embedding distance as an integral
    probability metric.
  \item \textbf{Information-theoretic family (UNPROVEN, easy).}
    Mutual information, transfer entropy, and effective information
    are all KL divergences against appropriate factored references.
    Schreiber's transfer entropy is one line; Hoel's EI is the
    intervention-averaged KL.
  \item \textbf{Unification (UNPROVEN, the load-bearing step).}
    Following Reid \& Williamson, all three reduce to a single
    schema modulo monotone reparametrisations. The exact equivalence
    statement depends on how strict the schema is required to be.
\end{enumerate}

The Agrawal--Horel optimal-bounds result on f-divergences vs.\ IPMs
suggests the right statement is ``equivalent up to monotone
reparametrisation'' rather than literal equality.

\section{Status and follow-up}

\textbf{Current status: UNPROVEN.} Plausibly the lowest-cost lemma to
prove. A positive C3 tightens the typing of $\mathbf{Obs}$ for
N1's Decomp construction.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
