% =============================================================================
%  S1: Sufficiency — skeleton paper
%  STATUS: UNPROVEN. This file is a skeleton; the proof is not attempted here.
% =============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem*{namedconjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\X}{\mathbf{X}}

\title{Sufficiency of the Four-Slot Framework for $\mathbf{LifeCat}$ Membership \\
       \large (UNPROVEN --- skeleton paper for S1)}
\author{Liam McGuire\thanks{Unaffiliated researcher, Seattle, USA. liammcg44@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper is a \textbf{skeleton} that records the statement of the
sufficiency conjecture S1 from the
scaffolding paper~\cite{emergent_systems_2026} and identifies
\texttt{l1\_causal\_emergence\_gol/} as the natural concrete witness.
\textbf{No proof is attempted in this paper as of its current
revision.} The result is UNPROVEN.
\end{abstract}

\section{Statement (UNPROVEN)}

\begin{quote}
\emph{``S1 (sufficiency): there exists a 4-tuple system $\mathcal{S}$ and
an external observer functor $O$ such that the pair $(\mathcal{S}, O)$
has vitality profile $(k, \boldsymbol{\sigma})$ with $k \geq 1$, hence
$(\mathcal{S}, O) \in \mathbf{LifeCat}$.''}
\end{quote}
\noindent (Scaffolding paper~\cite{emergent_systems_2026},
\S\texttt{sec:s1}.)

\section{Plan of attack (no proof attempted)}

Two natural routes.

\begin{enumerate}[leftmargin=2em,itemsep=4pt]
  \item \textbf{Witness route (UNPROVEN).} Cite the companion paper
    \texttt{l1\_causal\_emergence\_gol/} (also UNPROVEN). A positive L1
    result automatically yields a sufficiency witness for S1: the
    Game-of-Life-under-Markov-blanket system satisfies the three
    $\mathbf{LifeCat}$ conditions
    (Definition~\texttt{def:lifecat} of the scaffolding paper).
  \item \textbf{Abstract route (UNPROVEN).} Construct an
    $\mathcal{O}_W$-algebra in $\mathbf{Stoch}$ that satisfies the
    membership predicate by direct construction, without needing the
    causal-emergence verification to be carried out on a specific
    finite system.
\end{enumerate}

Either route discharges S1. The witness route is cheaper and is the
default plan. The abstract route is a fallback if L1 turns out to be
obstructed.

\section{Status and follow-up}

\textbf{Current status: UNPROVEN.} Subsumed by L1 in practice.
\texttt{RESULTS.md} should track S1 alongside L1 and flip them
together when the L1 witness lands.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
