% =============================================================================
%  Horizontal composition formalisation — skeleton paper
%  STATUS: UNPROVEN. Could be folded into N1 rather than written separately.
% =============================================================================
\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[expansion=false]{microtype}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\title{Horizontal Composition $\boxtimes_\kappa$ as the Monoidal Product \\
       of $\mathbf{Sys}$, with $\sigma$-Uplift Formula \\
       \large (UNPROVEN --- skeleton paper)}
\author{Liam McGuire\thanks{Unaffiliated researcher, Seattle, USA. liammcg44@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper is a \textbf{skeleton} for a formal treatment of
$\boxtimes_\kappa$ as the monoidal product of $\mathbf{Sys}$, with
the precise $\sigma$-uplift formula across the three coupling cases
(trivial, one-way, mutual). \textbf{No proof is attempted.} The paper
exists as an optional standalone target; the content may
alternatively be folded into \texttt{n1\_necessity/} when that paper
develops the monoidal structure of $\mathbf{Sys}$. The result is
UNPROVEN.
\end{abstract}

\section{Goal (UNPROVEN)}

Formal treatment of horizontal composition. Three main lemmas:

\begin{enumerate}[leftmargin=2em,itemsep=4pt]
  \item \textbf{$\boxtimes_\kappa$ is monoidal (UNPROVEN).} Verify
    the SMC axioms (associator, unitor, symmetry, hexagon, pentagon)
    on the substrate-composition operator. Most of this is in the
    scaffolding paper §3.3; making it precise in the SMC vocabulary
    is the work.
  \item \textbf{$\sigma$-uplift formula (UNPROVEN).} For two systems
    $\mathcal{S}_1, \mathcal{S}_2$ at common depth $k$ with profiles
    $(k, \boldsymbol{\sigma}_1)$ and $(k, \boldsymbol{\sigma}_2)$,
    the composite $\mathcal{S}_1 \boxtimes_\kappa \mathcal{S}_2$ has
    $\sigma_i$ at each level $i$ given by a formula in $\sigma_i^{(1)},
    \sigma_i^{(2)}$, and the coupling structure $\kappa$. The
    scaffolding paper §\texttt{sec:horizontal} gives the three cases
    informally; this paper makes them precise.
  \item \textbf{Three coupling cases (UNPROVEN).} Verify the
    saturation-at-1 behaviour for trivial / one-way / mutual
    coupling; in particular, that mutual coupling can manufacture
    $\sigma$ at the shared level.
\end{enumerate}

\section{Status and follow-up}

\textbf{Current status: UNPROVEN.} May be folded into
\texttt{n1\_necessity/}; alternatively published standalone for
operadic-theory audiences.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
